
---
title: "3rd and 6 Plays"
author: "Tahlia Williams"
date: '2025-10-22'
format: 
  html:
    code-fold: true
---

```{r}
library(tidyverse)
 library(lubridate)
```

## Close Rushing Wins: How Rare Are They?

It is normally rare for football teams to win without running the ball effectively, but through data it isnt impossible. 

Since the 2015 season, fewer than 60% of college football victories have come when the winning team rushed for under 100 yards.

```{r}
logs <- read_csv("https://dwillis.github.io/sports-data-files/footballlogs1525.csv")

# the Season is the year of the game, except when the game is in January. Then its year-1

logs <- logs |> mutate(Season = case_when(
  month(Date) == 1 ~ year(Date) - 1, 
  .default = year(Date)
))

logs |> select(Date, Season) 

season_games <- logs |> 
  group_by(Season) |> 
  summarise(all_games = n())


season_games_under_100_wins <- logs |> 
  filter(Result == "W", RushYds < 100) |> 
  group_by(Season) |> 
  summarise(low_rush_wins = n())

low_rush_wins <- season_games |> 
  inner_join(season_games_under_100_wins, join_by(Season)) |> 
  mutate(pct_low_rush_wins = low_rush_wins/all_games * 100)

bigten_logs <- logs |> filter(Conference == "Big Ten Conference")

bigten_season_games <- bigten_logs |> 
  group_by(Season) |> 
  summarise(all_games = n())


bigten_low_rush_wins <- bigten_logs |> 
  filter(Result == "W", RushYds < 100) |> 
  group_by(Season) |> 
  summarise(low_rush_wins = n())

bigten_combined <- left_join(bigten_season_games, bigten_low_rush_wins, by = "Season") |> 
  mutate(percentage = (low_rush_wins / all_games) * 100)

bigten_combined |> 
  ggplot(aes(x = Season, y = percentage)) +
  geom_line(linewidth = 1) +
  geom_point() +
  geom_line(data = bigten_combined, aes(x = Season, y = percentage), linewidth = 1) +
  geom_point(data = bigten_combined, aes(x = Season, y = percentage)) +
  labs(title = "Rare Rushing Wins: Under 100 Yards to Victory",
       x = "Season",
       y = "Percentage of Wins",
       color = "Group") +
  theme_minimal()

```

The chart above shows that low-rushing wins make up only a small percentage of all games, with Big Ten teams pulling them off less. The Big Ten is known for its physical, run-heavy style, so when a team wins with fewer than 100 rushing yards, they are going to stand out.

These wins show how the game is evolving toward more pass-first offenses and faster scoring drives, especially in programs willing to lean on their quarterbacks and receivers when the ground game stalls.

```{r}
team_results <- logs |>
group_by(Team) |>
summarise(
Wins = sum(Result == "W"),
Losses = sum(Result == "L")
) |>
arrange(desc(Wins))


season_team_results <- logs |>
group_by(Season, Team) |>
summarise(
Wins = sum(Result == "W"),
Losses = sum(Result == "L")
) |> arrange(Season, desc(Wins))


maryland_outcomes <- logs |>
filter(Team == "Maryland") |> 
group_by(Season, Team) |>
summarise(
Wins = sum(Result == "W"),
Losses = sum(Result == "L")
) |>
arrange(Season, desc(Wins))

bigten_results <- bigten_logs |>
group_by(Season, Team) |>
summarise(
Wins = sum(Result == "W"),
Losses = sum(Result == "L")
) |>
arrange(Season, desc(Wins))


maryland_results <- bigten_results |> filter(Team == "Maryland")




ggplot(maryland_outcomes, aes(x = Season)) +
geom_col(aes(y = Wins), fill = "blue") +
geom_col(aes(y = -Losses), fill = "red") +
labs(
title = "Maryland Football Wins and Losses by Season",
x = "Season", y = "Wins (above) / Losses (below)"
) +
theme_minimal()

```
While running the ball still matters, modern offenses prove that efficiency and balance—not just total yards—decide close games.



